Design principles of transcription factors with intrinsically disordered regions

Version 1

Abstract

Abstract

Transcription Factors (TFs) are proteins crucial for regulating gene expression. Effective regulation requires the TFs to rapidly bind to their correct target, enabling the cell to respond efficiently to stimuli such as nutrient availability or the presence of toxins. However, the search process is hindered by slow diffusive movement and the presence of ‘false’ targets - DNA segments that are similar to the true target. In eukaryotic cells, most TFs contain an Intrinsically Disordered Region (IDR), which is a long, flexible polymeric tail composed of hundreds of amino acids. Recent experimental findings indicate that the IDR of certain TFs plays a pivotal role in the search process. However, the principles underlying the IDR’s role remain unclear. Here, we reveal key design principles of the IDR related to TF binding affinity and search time. Our results demonstrate that the IDR significantly enhances both of these aspects. Furthermore, our model shows good agreement with experimental results, and we propose further experiments to validate the model’s predictions.

Transcription Factors (TFs) are fundamental proteins that regulate gene expression: they bind to specific DNA sequences to control gene transcription [1]; they respond adaptively to various cellular signals, thereby modulating gene expression [2, 3]; they guide processes such as cell differentiation and development, and their malfunctions are often linked to disease onset [4, 5]. The regulation mechanism is via the binding of a TF to the regulatory DNA regions associated with the genes, the TF’s target. Effective regulation requires TFs to have sufficient binding affinity and a binding rate comparable to cellular processes. In eukaryotic cells, besides having a DNA-binding domain (DBD), approximately 80% of TFs also have one or more intrinsically disordered regions (IDRs), which lack a stable 3D structure and can be considered as flexible, unstructured polymers, hundreds of amino acids (AAs) in length [6–9]. Molecular dynamics simulations suggest that IDRs play a significant role in promoting target recognition [10], by increasing the area of effective interaction between the TF and the DNA. Upon binding, the search problem becomes effectively a one-dimensional diffusion process [11]. Recent in vivo experiments done in yeast also suggest that IDRs could take a key part in guiding TFs to the DNA regions containing their targets [12, 13].

The problem of TF search has been widely studied, theoretically, albeit with the majority of works focusing on parameter regimes relevant for bacteria. One important mechanism is that of facilitated diffusion, in which the TF performs 1D diffusion along the DNA and at any given moment can fall off and perform 3D diffusion until reattaching to the DNA (a 3D excursion). The 1D diffusion along the DNA and the 3D excursions alternate until the TF finds its target. This mechanism is well characterized theoretically [14–20] and supported by experiments in E. coli [21]. However, due to the significant structural differences, including larger genome size and chromatin structure, the facilitated diffusion mechanism cannot explain the fast search times observed in eukaryotes, and a qualitatively different mechanism is required [13, 22–25].

Here, we attempt to address the fundamental question of how IDRs can enhance the binding affinity of the TFs and speed up their search process, and to discover the associated design principles in eukaryotes. The problem of binding affinity (binding probability) exhibits the classical trade-off between energy and entropy: on one hand, the binding of IDR to the DNA is energetically favorable. On the other hand, this binding leads to constraints on the positions of the IDR (pinned to the DNA at the points of attachment) which leads to a reduction in the entropy associated with the possible polymer conformations - hence an increase in the free energy of the system. Therefore the effects of the disordered tail on affinity are non-trivial, leading to physical problems reminiscent of those associated with the well-studied problem of DNA unzipping [26]. We find that the binding probability increases dramatically for TFs having IDRs (see Fig. 1b) for a broad parameter regime. The optimal search process consists of one round of3D-1D search, where at first the TF performs 3D diffusion and then binds the antenna and performs 1D diffusion until finding its target, rarely unbinding for another 3D excursion. This process is presented in Fig. 1(a, c). We obtain an expression for the search time that shows good agreement with our numerical results.

Model illustration of a TF with the IDR and its search process.

a, Illustration of a TF locating its target site (highlighted in orange) on the DNA strand (depicted as a green curve). Surrounding the target is a region of length L, where the TF’s IDR interacts with the DNA. The TF performs 3D diffusion until it encounters and binds to this “antenna” region. b, Illustration of a TF that includes an IDR composed of AAs. On the right: a model of the TF, a polymer chain, comprising of a single binding site on the DNA Binding Domain (DBD) and multiple binding sites on the IDR. c, Once bound to the antenna, the TF performs effective 1D diffusion until it reaches its target. The 1D diffusion is via the binding and unbinding of sites along the IDR, a process we coin “octopusing”.

Results

Binding probability of the TF from the equilibrium approach

The experiment described in Ref. [12] studied the binding affinity of wild-type and mutated TFs to their corresponding targets. They found that shortening the IDR length results in weaker affinity. To qualitatively model the binding affinity of a TF to its target, we utilize a well-known physical model of polymers, the Rouse Model [27, 28] (used in the subsequent dynamical section as well). See Materials and Methods. We note that our conclusions do not hinge on this choice, and our results hold also when using other polymer models (see Supplementary S1). Since the IDR is thought to recognize a relatively short flanking DNA sequence of the DBD target [12], we model a DNA segment as a straight-line “antenna” region [17, 29–31], where the IDR targets reside. We assume that the interaction between TF and DNA outside the antenna region is negligible. Note that the interaction between IDR and DNA is a specific interaction, with genomic localization determined by multivalent interactions mediated by hydrophobic residues [32, 33].

To gain an intuition into the binding probability of TF, we first study a TF with a single IDR site. From here on, energies will be measured in units of kBT , to simplify our notation. We assume that the IDR has only one corresponding target. There are four possible states: neither the DBD nor the IDR are bound, the IDR is bound, the DBD is bound, and both are bound, which correspond to Pfree, PIDR, PDBD , and Pboth, respectively. The probability that the TF is bound at the DBD target is the sum of PDBD and Pboth. In thermal equilibrium, each of the 4 terms is given by the Boltzmann distribution, which can be solved analytically by integrating-out the additional degrees of freedom associated with the polymer (Supplementary S2). To better understand the associated design principles, it will be helpful to compare this probability with that of a “simple” TF (i.e., without the IDR), which we denote by Psimple. We find that: PTFPsimple≈(1+P*), (1)

where 𝒫 is an enhancement factor, governing the degree to which the IDR improved the binding affinity. Importantly, we find that: P*(EB,d,l0)=eEBV1f(d,l0), (2)

where f(d,l0)=(32πl02)3/2exp(3d22l02) , V1 is the target volume, −EB is the binding energy for the IDR target (EB is always positive), d is the distance between the DBD target and the IDR target, and l0 is the distance between the DBD and the IDR binding site on the TF (l0 and d are also the distances for neighboring IDR targets and neighboring IDR sites, respectively, in the case of multiple IDR targets and sites). This shows that the presence of the IDR always increases the binding affinity since 𝒫 is always positive. However, the magnitude of enhancement can be negligible: For instance, even if EB is large, the term f(d, l0), associated with the entropic penalty (because the end-to-end distance is fixed, reducing the degrees of freedom compared to a free polymer), could lead to small values of 𝒫. The exponential decay of 𝒫 with d2 (through the term f(d, l0)), indicates that placing the IDR targets close to the DBD target is preferable. However, due to physical limitations, d cannot be arbitrarily small and in reality will have a lower bound. Once d is fixed, 𝒫 is maximal at l0 = d, as can be gleaned from equation (2). This condition minimizes the free energy penalty associated with the entropic contribution of the term f(d, l0). As we shall see, this design principle is also intact when multiple IDR sites and targets are involved, which is the scenario we discuss next.

A priori, it is unclear whether having more binding sites (nb > 1) on the IDR or more targets (nt > 1) on the DNA will be beneficial or not. We therefore sought to investigate how the binding affinity depends on these parameters, as well as the binding energy and positions of the targets and binding sites. This generalization requires taking into account all the possible number of bindings, n (ranging from 0 to min (nb, nt)), and all possibilities for binding orientations for each n, (nbn)(ntn)n! distinct configurations in total. For each of them, we may analytically integrate-out the polymer degrees of freedom and calculate explicitly the contribution to the partition function (Supplementary S2). Summing these terms together, numerically, allows us to explore how the different parameters affect the binding specificity, which we describe next. To characterize the binding affinity, we resort again to the ratio Q = PTF/ Psimple, measuring the effectiveness of the binding in comparison to the case with no IDRs. The length of TFs in our considerations, which varies with nb, is comparable to that in experiments [12].

We find the design principle: l0 = d, identical to what we showed for the case of a single IDR binding site. Specifically, Q is maximal at l0 ~ d for all combinations of nb and nt. Fig. 2a displays this principle for several combinations of nb and nt. All combinations are provided in Supplementary Fig. S3. This is plausible since for the IDR to be bound to the DNA at multiple places without paying a large entropic penalty, the distance between target sites should be compatible with the distance between binding sites. We therefore set l0 = d and search for other design principles. Fig. 2b shows a heatmap of Q varying with nb and nt. At large nt , as nb increases, the value of PTF exhibits an initial substantial increase, followed by a gradual rise until it reaches saturation (shown explicitly in Supplementary Fig. S4). A similar behavior is also observed when varying the value of nt. (When nt exceeds nb , the contribution from the states where the IDR is bound but the DBD is unbound increases with increasing nt, which slightly lowers PTF; see Supplementary Fig. S4.) In fact, the heatmap shown in Fig. 2b shows that the value of PTF is governed, approximately, by the minimum of nb and nt (this is reflected in the nearly symmetrical “L-shapes” of the contours). This places strong constraints on the optimization, and suggests that there will be little advantage in increasing the value of either nb or nt beyond the point nb = nt , revealing another design principle (under the assumption that it is preferable to have smaller values of nb and nt for a given level of affinity, in order to minimize the cellular costs involved). We verify this principle holds at additional values of EB. Fig. 2c shows that the contours corresponding to PTF = 0.9 at varying values of EB all take an approximate symmetrical L-shape, thus supporting the design principle nb = nt. For a given EB, we denote the optimal values as nb*=nt* , and show them as hexagrams in the plot. We further find that nb* decreases with increasing EB, as displayed in the inset. These relationships also hold at PTF = 0.5 (Supplementary S5).

We examine the variation of Q with respect to EB while keeping nb = nt constant. In Fig. 2d, as the binding energy EB increases, Q increases from a value of ~ 1 to its saturation value, 1/Psimple, in a switch-like manner: i.e., there is a characteristic energy scale, Eth. To estimate Eth, when Q ≪ 1/Psimple (i.e., PTF ≪ 1), we approximate the expression of Q as: Q≈1+∑n=1nb∑orn∏i=1nP*(EB,rorn,i,lorn,i), (3)

where lorn,i and rorn,i denote the segment length between two adjacent bound sites and the distance associated with their respective targets, respectively, and “orn” represents the orientation of the given configuration (Supplementary S2). Equation (3) reduces to equation (1) when nt = nb = 1. To substantially increase Q, the configuration that contributes the most to Q, would satisfy (𝒫(EB, d, d))nb ≫ 1. Thus, the characteristic energy Eth could be estimated using the formula 𝒫(Eth, d, d) = 1, resulting in: Eth=32[ln(2π3ϕ2)+1], (4)

where ϕ=V11/3/d represents the proportion of IDR targets within the antenna. Equation (4) is a direct consequence of the energy-entropy trade-off. Increasing ϕ leads to a decrease in entropy loss due to binding the IDR, subsequently reducing the demands on binding energy. Given that ϕ ≪ 1, Eth is much greater than 1kBT. The vertical dashed line in Fig. 2d, obtained from equation (4), effectively captures the region of significant increase for large values of nb. Hence, we obtain the third design principle: EB > Eth, which ensures the significant binding advantage of TFs with IDRs.

Design principles across various parameters: l0/d, nb, nt, and EB.

a, Plots of Q = PTF/Psimple varying with l0/d are shown for several combinations of nt and nb for typical values of the relevant quantities: the nucleus volume is 1μm3, EDBD = 15, EB = 10, V1 = (0.34nm)3, and d=50⋅0.34nm . b, A heatmap of Q vs. nb and nt at EB = 10 and d = l0. c, Contour lines at different EB at PTF = 0.9. Blue hexagrams represent the design principle nb*=nt* , with the black dashed line as a visual guide. The inset shows the dependence of nb* on energy. d, Q for varying EB at d = l0 for several nb = nt. The vertical line represents Eth as determined by equation (4), the solid curves illustrate the approximation of Q obtained from equation (3), and the horizontal line indicates where PTF = 1.

So far, we have found that the binding design principles for TFs with IDRs are as follows: (i) target distance d should be as small as possible, (ii) IDR segment binding distance l0 should be comparable to d, (iii) itis preferable to have fewer binding sites nb* and binding targets nt* as the binding strength EB increases, with nb*∼nt* , and (iv) EB should be larger than a threshold that solely depends on the proportion ϕ of IDR targets within the antenna, as indicated by Eq. (4). In our estimation, with ϕ2 = 0.02, we obtain Eth ≈ 8.5kBT. The above principles have also been validated in scenarios involving Poisson statistics of binding site distances and target distances, thus indicating their robust applicability (Supplementary Fig. S6).

Search time of the TF from the dynamics

In this section, we study the design principles of the TF search time, a dynamic process. Here, we confine the TF with ñ = nb+1 sites (nb IDR sites and one DBD site) to a sphere of radius R, the cell nucleus, and use over-damped coarse-grained molecular dynamics (MD) simulations. The sites and targets interact via a short-range interaction. The DBD target with a radius a is placed at the sphere’s center. See the sketch in Fig. 1a. Once the DBD finds its target, the search ends. We denote the mean total search time (i.e., the mean first passage time, MFPT), as ttotal. We estimate itby averaging 500 simulations, each starting from an equilibrium configuration of the TF at a random position in the nucleus. Simulation details are provided in the Materials and Methods section.

By conducting comprehensive simulations of ttotal across various parameter values of L, EB, and ñ, we were able to identify its minimum value, which is approximately at L* = 300bp, EB* = 11, and ñ* = 4 for R = 500bp ≈ 1/6μm. Here, one base pair (bp) is equal to 0.34nm. To speed up the simulations, we utilized several computational tricks, including adaptive time steps and treating the TF as a point particle when it is far from the antenna (see Materials and Methods). Despite of this, the simulations require 105 CPU hours due to the IDR trapped in the binding targets on the antenna, which results in the slow dynamics as found in the previous simulation works [10, 34]. The probability distribution function (pdf) of individual search times exponentially decays with a typical time equal to ttotal, see inset of Fig. 3. This indicates that ttotal provides a good characterization of the search time.

We find that ttotal varies non-monotonically with L at fixed EB* and ñ* as shown in Fig. 3. Once a TF attaches to one of its targets, it will most likely diffuse along the antenna until reaching the DBD target (Supplementary S7). It means that the optimal search process is a single round of 3D diffusion followed by 1D diffusion along the antenna. In Supplementary S8, by analytically calculating the mean total search time using a coarse-grained model, we find that a single round of 3D-1D search is typically the case in the optimal search process. Note that our model differs profoundly from the conventional facilitated diffusion (FD) model in bacteria: In the latter, the TF is assumed to bind non-specifically to the DNA. Within our model, the antenna length L can be tuned by changing the placement of the IDR targets on the DNA, making it an adjustable parameter rather than the entire genome length. As opposed to the FD model, which requires numerous rounds of DNA binding-unbinding to achieve minimal search time, within our model, the minimal search time is achieved for a single round of 3D and 1D search.

Interestingly, when we observe the motion of a TF along the antenna, IDR sites behave like “tentacles”, as illustrated in Fig. 1c, constantly binding and unbinding from the DNA. We refer to this 1D walk as “octopusing”, inspired by the famous “reptation” motion in entangled polymers suggested by de Gennes [35] (see Supplementary Video). We note that previous numerical studies [10, 11] used comprehensive, coarse-grained, MD simulations to demonstrate that including heterogeneously distributed binding sites on the IDR could facilitate intersegmental transfer between different regions of the DNA, which is reminiscent of our octopusing scenario.

The mean 3D search time, t3D, and the mean 1D search time, t1D are also shown in Fig. 3. To evaluate t3D, we take advantage of several insights: first, if the TF gyration radius, rp=n˜/6l0 , is comparable or larger than the distance between targets, d, we may consider the chain of targets as, effectively, a continuous antenna. Making the targets denser would hardly accelerate the search time, reminiscent of the well-known chemoreception problem where high efficiency is achieved even with low receptor coverage [36]. Second, when the distance between the TF and the antenna is smaller than rp, one of the binding sites on the TF would almost surely attach to the targets. This allows us to approximately map the problem to the MFPT of a particle hitting an extremely thin ellipsoid with long-axis of length L and radius rp, the solution of which is ln(L/rp)/L [37]. We find that this scaling captures well the dependence of t3D on the parameters, with a prefactor α ~ O(1). For the 1D search from a single particle picture, since the TF is unlikely to detach once it attaches, it performs a 1D random walk. So, taken together, the mean total search time is ttotal=t3D+t1D≈αn˜R3DLln(L/rp)+βL2D, (5)

where D is the 3D diffusion coefficient of one site. For a TF with ñ sites, its 3D diffusion coefficient, D3, is D/ñ. α and β are fitting parameters. The semi-analytical formulas of t1D, t3D, and ttotal, denoted by the solid lines in Fig. 3, agree well with the simulation results for α = 0.5 and β = 14.5. The theoretical result for the time spent in 3D was originally calculated for a point searcher [37]. In that case, a value of α = 2/3 was obtained. β is larger at larger EB (see Supplementary S7).

We also examine the dependence of ttotal on the key parameters EB and ñ, in the vicinity of the optimal parameters which minimize the total search time (Supplementary S7). We find that the affinity of TFs to the antenna decreases rapidly with decreasing the IDR length [12] and EB. Consequently, when ñ < ñ* or EB<EB* , multiple search rounds are needed to reach the DBD target, resulting in a significant escalation of ttotal. When ñ > ñ*, ttotal, t3D, and t1D agree with equation (5). When EB>EB* , since t3D is independent of EB, the variation in ttotal stems from t1D which exponentially increases with EB.

Mean search time varying with the antenna length L.

ttotal, t3D and t1D vs. L at EB* and ñ*. The solid curves correspond to ttotal , t3D and t1D in equation (5). Inset: probability density function (pdf) exponentially decays at large t. pdf ≈ exp(-t/ttotal)/ttotal (black solid line). The unit t0 represents the time for one AA to diffuse 1bp. a = 0.5bp, d= 10bp, and l0 = 5bp.

The minimal search time for a simple TF is tsimple = R3/(3Da) [37]. For the TF with an IDR, by minimizing the expression in equation (5), we find that the optimal length approximately scales as L* ∝ R, and therefore the minimal search time, tmin, is approximately proportional to R2. At R = 500 ≈ 1/6μm, the search time is more than ten times shorter than a simple TF, despite the latter’s diffusion coefficient being several times larger. Based on our analytic result, we expect ~ 50 fold enhancement for the experimental parameters of yeast (note that this formula is approximate since it does not take into account the persistence length of DNA [38], which is severalfold shorter than the optimal antenna length L*. We have verified that the mean total search time does not change significantly for a very curved antenna, as detailed in Supplementary S9.). We also verified that the 3D diffusion coefficient slightly changes under a complex DNA configuration, where a point-searcher TF non-specifically interacts with the entire DNA, indicating the robustness of our optimal search process (see Supplementary S10).

Next, we characterize the dynamics of the octopusing walk (Supplementary S11). During this process, different IDR binding sites constantly bind and unbind from the antenna, leading to an effective 1D diffusion. A priori, one may think that the TF faces conflicting demands: it has to rapidly diffuse along the DNA while also maintaining strong enough binding to prevent detaching from the DNA too soon. Indeed, within the context of TF search in bacteria, this is known as the speed-stability paradox [16]. Importantly, the simple picture of the octopusing walk explains an elegant mechanism to bypass this paradox, achieving a high diffusion coefficient with a low rate of detaching from the DNA. This comes about due to the compensatory nature of the dynamics: there is, in effect, a feedback mechanism whereby if only a few sites are bound to the DNA, the on-rate increases, and similarly, if too many are bound (thus prohibiting the 1D diffusion) the off-rate increases. Remarkably, this mechanism remains highly effective even when the typical number of binding sites is as low as two. Further details are provided in Supplementary S11.

Comparing with experimental results

In Ref. [12], the binding probability of the TF Msn2 in yeast was measured, where the native IDR was shortened by truncating an increasingly large portion of it. The results indicate that binding probability decreases with increasing truncation of the IDR, as illustrated in Fig. 4(a). We utilized our model to calculate the relative binding probability compared to PTF at a zero truncating length. Using the DBD binding energy EDBD as our only fitting parameter, our model achieves quantitative agreement with the experimental data. In Fig. 4(b), we estimate the search time as a function of L for R = 1μm and D0 = 5 × 10−10m2/s for a single AA. The estimated timescales are comparable to empirical results, which are 1 to 10 seconds [13, 39], as indicated by the shaded area. We have also estimated the on-rate kon and off-rate koff of the TF as they vary with IDR length, as shown in Fig. 4(c). In the kinetic proofreading mechanism [40], specificity is associated with the off-rate, but recent studies [41] suggest that in other scenarios, specificity is encoded in the on-rate. Therefore, it is important to explore how specificity is encoded within this mechanism. Our results indicate that the variation in the dissociation constant koff /kon is dominated by koff, suggesting that specificity is encoded in the off-rate. The magnitudes and variations of these rates could be directly compared to experimental results in future studies.

(a) Relative binding probability varies with the truncation length of the IDR, which quantitatively agrees with the experimental data in Ref. [12], where we only vary EDBD (binding energy of DBD) to ensure that the maximal value reaches 1 at zero truncation length. Antenna length L = 1000nm, EDBD = 22 and EB = 11. Other parameters are set the same as in Fig. 2. (b) The search time estimated by our model (from Eq. (5)) is quantitatively comparable with the experimental results, as guided by the shaded area. (c) The on-rate kon = ttotal/Vc and off-rate koff = (1 − PTF) / (PTFttotal) varying with the IDR length, indicating that kD = koff/kon ranges from 0.01 to 10 nM, with its variation primarily dominated by the off-rate.

In summary, we addressed the design principles of TFs containing IDRs, with respect to binding affinity and search time. We find that both characteristics are significantly enhanced compared to simple TFs lacking IDRs.

The simplified model we presented provides general design principles for the search process and testable predictions. One possible experiment that could shed light on one of our key assumptions is “scrambling” the DNA (randomizing the order of nucleotides). Our results depend greatly on our assumption that there exists a region on the DNA, close to the DBD target, that contains targets for the IDR binding sites. It is possible to shuffle, or even remove, regions of different lengths around the DBD target and to measure the resulting changes in binding affinity. According to our results, we would expect that shuffling a region smaller in length than the IDR should have minimal effect on the binding affinity and search time, but that shuffling a region longer than the IDR should have a noticeable effect.

Another useful test of our model would be to probe the dynamics of the TF search process by varying the waiting time before TF-DNA interactions are measured: Within the ChEC-seq technique [42] employed in Ref. [12], an enzyme capable of cutting the DNA is attached to the TF, and activated at will (using an external calcium signal). Sequencing then informs us of the places on the DNA at which the TF was bound at the point of activation. Measuring the binding probabilities at various waiting times — defined as the time between adding calcium ions and adding transcription factors, ranging from seconds to an hour, would inform us both of the TF search dynamics as well as the equilibrium properties: Shorter time scales would capture the dynamics of the search process, while longer time scales would reveal the equilibrium properties of the binding. At short waiting times (within several seconds), we would expect the binding probability to increase with waiting times. By analyzing a normalized binding probability curve as a function of the waiting time t, we could fit the curve using the function 1−exp(−t/tsearch). From this curve, we can infer the search time. Our results suggest that the presence of the IDR in a TF leads to increased binding probabilities over various waiting times and a reduced waiting time to reach a higher saturation binding probability. Furthermore, as shown in Fig. 4 obtained from our model, kon can be determined from the measurements of tsearch. Using the binding probability, which is associated with kD = koff /kon, koff can also be calculated.

While our work primarily focused on coarse-grained modeling of the search process and the hypothesized octopusing dynamics, higher-resolution molecular dynamics simulations could provide more detailed molecular insights and further validate our proposed design principles [10, 11]. Since these principles are grounded in fundamental physical concepts, such as the energy-entropy trade-off, they are likely to be robust across diverse molecular systems. We anticipate their applicability to other biophysical contexts, such as nonspecific RNA polymerase binding [43] and multivalent antibody binding [44], where the associated dynamic processes would also be of significant interest for future research.

Materials and methods

Model: The Rouse Model describes a polymer as a chain of point masses connected by springs and characterized by a statistical segment length l0 resulting from the rigidity of springs k and the thermal noise level kBT , l02 = 3kBT/k [27, 28]. We coarse-grain the TF by representing it with a few binding sites for the IDR, and an additional site for the DBD, as illustrated in Fig. 1b. The targets are assumed to be positioned at equal spacing, with one target for the DBD and a few targets for the IDR, corresponding to the orange and black squares in Fig. 1c. We have also verified that relaxing this assumption, and using randomly positioned targets, does not affect our results (Supplementary S6). In the Rouse Model, the 3D probability density describing the relative distance in the equilibrium state, denoted as r, between any two sites can be derived through Gaussian integration [28], leading to: f(r,l)=(32πl2)32exp(−3r22l2), (6)

where l=nsegl0 and nseg is the number of segments between the two sites.

Dynamical simulations: Each binding site of the TF at position ri follows over-damped dynamics, which reads: Δri=DkBTf({ri})Δt+2DΔtξ, (7)

where the random variable ξ follows the standard normal distribution. D is the diffusion coefficient of the coarsegrained binding site. The force acting on each site is f ({ri}) = −∂Utotal/∂ri, and Utotal=∑i=0nb−112k(ri+1−ri)2+∑i,jEBexp[−(ri−Rj)22wd2], (8)

where Rj are the positions of the targets. wd = 1bp is the typical width of the short-range interaction between sites and their corresponding targets. The typical distance l0 between coarse-grained binding sites is 5bp. The stiffness of the springs is thus 3kBT/l02=3kBT/25bp2 . To speed up the simulations, when the sites’ minimal distance to the antenna is larger than the threshold 10bp, the TF is regarded as a rigid body, and all sites move in sync with their center of mass. We verified that the resulting search time is insensitive to the value of this threshold. When the sites’ minimal distance to the antenna is less than 10bp, we consider the excluded volume effect to ensure that different binding sites are not trapped at the same target. Specifically, we use the WCA interaction [45]: UWCA={σ12|ri−rj|12−2σ6|ri−rj|6+1, |ri−rj|<σ0, |ri−rj|≥σ (9)

where σ = 1bp. In this region, we also adopted an adaptive time step: Δt/(bp2/D)=min{1,kBTmax|fi|bp}×0.01 . Note that this choice ensures that the displacement of the sites at every step is much smaller than the binding site size wd. Due to the IDR trapped in the binding targets on the antenna, which results in slow dynamics, the simulation time becomes significantly longer, as found in previous simulation studies [10, 34]. This slowdown is similar to the behavior observed in glass-forming liquids [46]. As a result, the individual simulation time steps range from 109 to 1011.

eLife Assessment

This paper present an important theoretical exploration of how a flexible protein domain with multiple DNA binding sites may simultaneously provide stability to the DNA-bound state and enables exploration of the DNA strand. The authors present compelling evidence that their findings have implications for the way intrinsically disordered regions (IDR) of transcription factors proteins (TF) can enhance their ability to efficiently find their binding site on the DNA from which they exert control over the transcription of their target gene. The paper concludes with a comparison of model predictions with experimental data which gives further support to the proposed model.

Reviewer #1 (Public review):

Summary:

The authors define the principles that, based on first principles, should be guiding the optimisation of trascription factors with intrinsically disordered regions (IDR). The first part of the study defines the following principles to optimize the binding affinities to the genome in the receiving region that is called the "antenna": (i) reduce the target to IDR-binding distance on the genome, (ii) optimise the distance betwee the DNA binding domain and the binding sites on the IDR to be as close as possible to the distance between their binding sites on the genome; (iii) keep the same number of binding sites and their targets and modulate this number with binding strength, reducing them with increased strenght; (iv) modulate the binding strenght to be above a threshold that depends on the proportion of IDR binding sites in the antenna. The second part defines the scaling of the seach time in function of key parameters such as the volume of the nucleus, and the size of the antenna, derived as a combination of 3D search of the antenna and 1D "octopusing" on the antenna. The third part focuses on validation, where the current results are compared to binding probabilith data from a single experiment, and new experiment are proposed to further validate the model as well as testing designed transcription factors.

Strengths:

The strength of this work is that it provides simple, interpretable and testable theoretical conclusions. This will allow the derived design principles to be understood, evaluated and improved in the future. The theoretical derivations are rigorous. The authors provides a comparison to experiments, and also propose new experiments to be performed in the future, this is a great value in the paper since it will set the stage and inspire new experimental techniques. Further, the field needs inspiration and motivations to develop these techniques, since they are required to benchmark the transcription factors designed with the methods presented in this paper, as well as to develop novel data based or in vivo methods that would greatly benefit the field. As such, this paper is a fundamental contribution to the field.

Weaknesses:

The model assumption that the interaction between the transcription factor and the DNA outside of the antenna region is negligible is probably too strong for many/most transcription factors, particularly in organisms with a longer genome than yeasts. The model presents many first principles to drive the design of transcription factor, but arguably, other principles and mechanisms might also play a role by being beneficial to the search and binding process. Specifically: (i) a role of the IDR in complex formation and cooperativity between multiple trascription factors, (ii) ability of the IDR to do parallel searching based on multiple DNA binding sites spaced by disordered regions, (iii) affinity of the IDR to specific compartmentalisations in the nucleus reducing the search time, etc. The paper would be improved by a discussion over alternative mechanisms.

Reviewer #2 (Public review):

Summary:

This is an interesting theoretical exploration of how a flexible protein domain, which has multiple DNA-binding sites along it, affects the stability of the protein-DNA complex. It proposes a mechanism ("octopusing") for protein doing a random walk while bound to DNA which simultaneously enables exploration of the DNA strand and stability of the bound state.

Strengths:

Stability of the protein-DNA bound state and the ability of the protein to perform 1d diffusion along the DNA are two properties of a transcription factor that are usually seen as being in opposition of each other. The octopusing mechanism is an elegant resolution of the puzzle of how both could be accommodated. This mechanism has interesting biological implications for the functional role of intrinsically disordered domains in transcription factor (TF) proteins. They show theoretically how these domains, if flexible and able to make multiple weak contacts with the DNA, can enhance the ability of the TF to efficiently find their binding site on the DNA from which they exert control over the transcription of their target gene. The paper concludes with a comparison of model predictions with experimental data which gives further support to the proposed model. Overall, this is an interesting and well executed theoretical paper that proposes an interesting idea about the functional role for IDR domains in TFs.

Weaknesses:

IDR domains are assumed flexible which I believe is not always the case. Also, I'm not sure how ubiquitous are the assumed binding sites on the DNA for multiple subdomains along the IDR. These assumptions though seem like interesting points of departure for further experiments.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Public review):

Summary:

The authors define the principles that, based on first principles, should be guiding the optimisation of trascription factors with intrinsically disordered regions (IDR). The first part of the study defines the following principles to optimize the binding affinities to the genome in the receiving region that is called the ”antenna”: (i) reduce the target to IDR-binding distance on the genome, (ii) optimise the distance betwee the DNA binding domain and the binding sites on the IDR to be as close as possible to the distance between their binding sites on the genome; (iii) keep the same number of binding sites and their targets and modulate this number with binding strength, reducing them with increased strenght; (iv) modulate the binding strenght to be above a threshold that depends on the proportion of IDR binding sites in the antenna. The second part defines the scaling of the seach time in function of key parameters such as the volume of the nucleus, and the size of the antenna, derived as a combination of 3D search of the antenna and 1D ”octopusing” on the antenna. The third part focuses on validation, where the current results are compared to binding probabilith data from a single experiment, and new experiment are proposed to further validate the model as well as testing designed transcription factors.

Strengths:

The strength of this work is that it provides simple, interpretable and testable theoretical conclusions. This will allow the derived design principles to be understood, evaluated and improved in the future. The theoretical derivations are rigorous. The authors provides a comparison to experiments, and also propose new experiments to be performed in the future, this is a great value in the paper since it will set the stage and inspire new experimental techniques. Further, the field needs inspiration and motivations to develop these techniques, since they are required to benchmark the transcription factors designed with the methods presented in this paper, as well as to develop novel data based or in vivo methods that would greatly benefit the field. As such, this paper is a fundamental contribution to the field.

Weaknesses:

The model assumption that the interaction between the transcription factor and the DNA outside of the antenna region is negligible is probably too strong for many/most transcription factors, particularly in organisms with a longer genome than yeasts. The model presents many first principles to drive the design of transcription factor, but arguably, other principles and mechanisms might also play a role by being beneficial to the search and binding process. Specifically: (i) a role of the IDR in complex formation and cooperativity between multiple trascription factors, (ii) ability of the IDR to do parallel searching based on multiple DNA binding sites spaced by disordered regions, (iii) affinity of the IDR to specific compartmentalisations in the nucleus reducing the search time, etc. The paper would be improved by a discussion over alternative mechanisms.

We thank the reviewer for highlighting that our work delivers simple, interpretable and rigorously derived conclusions, backed by experimental comparison and concrete proposals for future studies.

Regarding interactions outside the antenna region, Supplementary S10 shows that the non-specific IDR–DNA interactions (on the order of 1 kBT) only slightly alter the 3D diffusion coefficient and thus do not affect our conclusions regarding the optimal search process.

We have also added sentences in the discussion section regarding the alternative mechanism.

Reviewer #2 (Public review):

Summary:

This is an interesting theoretical exploration of how a flexible protein domain, which has multiple DNAbinding sites along it, affects the stability of the protein-DNA complex. It proposes a mechanism (”octopusing”) for protein doing a random walk while bound to DNA which simultaneously enables exploration of the DNA strand and stability of the bound state.

Strengths:

Stability of the protein-DNA bound state and the ability of the protein to perform 1d diffusion along the DNA are two properties of a transcription factor that are usually seen as being in opposition of each other. The octopusing mechanism is an elegant resolution of the puzzle of how both could be accommodated. This mechanism has interesting biological implications for the functional role of intrinsically disordered domains in transcription factor (TF) proteins. They show theoretically how these domains, if flexible and able to make multiple weak contacts with the DNA, can enhance the ability of the TF to efficiently find their binding site on the DNA from which they exert control over the transcription of their target gene. The paper concludes with a comparison of model predictions with experimental data which gives further support to the proposed model. Overall, this is an interesting and well executed theoretical paper that proposes an interesting idea about the functional role for IDR domains in TFs.

Weaknesses:

IDR domains are assumed flexible which I believe is not always the case. Also, I’m not sure how ubiquitous are the assumed binding sites on the DNA for multiple subdomains along the IDR. These assumptions though seem like interesting points of departure for further experiments.

We thank the reviewer for their careful and insightful evaluation of our work. In particular, we appreciate your emphasis on the inherent trade-off between binding stability and one-dimensional diffusion, and your recognition of how the octopusing mechanism elegantly reconciles these conflicting requirements.

To address the flexibility of TFs with IDRs, we incorporated the spring’s rest length—effectively introducing tunable rigidity—in Supplementary Section S1, and we show that our design principles for binding probability remain robust. Indeed, this is a highly interesting point; a comprehensive study will require more detailed modeling alongside experimental validation.

We acknowledge that the current evidence for IDR-directed DNA binding is primarily derived from a limited number of well-studied cases, particularly Msn2 in yeast, and the ubiquity of this mechanism across diverse transcription factors remains to be established.

Reviewer #1 (Recommendations for the authors):

The paper jumps to fast to the results, an larger introduction might improve the paper, the current introduction jumps too fast to results. Further, line 50, I don’t think that the figure is properly referenced. The formula 2 is confusing since what is the target volume V1 is not explained in the context of the formula, please expand the explanations.

We appreciate the reviewer’s valuable recommendations. We have expanded the Introduction, clarified V1, and updated the line 50.

Reviewer #2 (Recommendations for the authors):

I have some mostly minor suggestions to the authors for improving the manuscript:

In the abstract and introduction on at least two occasions the authors talk about IDRs as though they’re necessarily flexible. My understanding is that, while this is a very reasonable assumption, I don’t think this is something we know with any certainty for most IDRs. If the authors agree with my assessment I think they should reflect this uncertainty in the writing.

Thank you for the recommendations. We revised the wording to reflect the uncertainty, changing it to: “... commonly assumed to behave as a long, flexible...” and “...can be assumed as flexible....”.

It took me a bit of time to figure out what’s going on in Figure 1b. To help the reader I would suggest labeling the DBD targets (yellow square) and the IDR targets (gray squares) as such. The figure also left me guessing whether the DBD domain can bind to the IDR targets non-specifically? (I presume not.) This also brought a slightly bigger question into focus for me, wouldn’t the presence of the IDR binding ”sites” (since these ”sites” are on the protein I think the term ”domains” instead of ”sites” ) mean that this would increase the time the protein is bound non-specifically somewhere far from the target thereby increasing the search time. Or is the ability of the protein to bind specifically to DNA away from the DBD target ignored?

We have labeled the DBD targets and IDR targets in the figure. ‘Domains’ usually refers to structured parts; we keep using ‘sites’ and clarify that they correspond to short linear motifs.

The reviewer is correct. Our model omits any non-specific binding between the DBD and IDR-binding targets, as well as between the TF and other DNA regions. If such interactions were to substantially lengthen the search time, they would effectively revert our mechanism to the classical bacterial facilitateddiffusion model, which is generally considered inappropriate for IDR-mediated TF search in eukaryotic cells. However, Supplementary Figure S10 demonstrates that non-specific IDR–DNA interactions induce only marginal changes in the effective three-dimensional diffusion coefficient within complex chromatin environments, and therefore do not alter our conclusions regarding the optimal search process.

In Equation 2 and the text that follows I was left wondering what is the target volume V1. Also, I think it would be helpful to the reader to give them a sense of scale for the dimension full quantities appearing in Equation 2. This is done later when comparing the theory to experimental data, but I think it would be helpful to give a sense of size earlier in the manuscript.

V1 denotes the volume of the IDR–binding target region, which is on the order of bp3. f(d,l0) has units of inverse volume. We have included the units and specified the order of magnitude of V1 after Equation 2.

The binding energy EB is discussed a number of times but it wasn’t clear to me that this quantity referred to the energy per IDR site on the DNA or the total energy when the IDR is bound to DNA. In Figure 1 it would seem that the model allows only one IDR domain bound at a given time but I think the model allows for multiple IDR domains to be bound to the IDR target sites simultaneously. Right? Maybe make this clear in the Figure and the text.

EB denotes the binding energy per binding site, where each site corresponds to a short linear motif. Yes, we allow for multiple IDR domains to be bound to the IDR target sites simultaneously. We have clarified the definition of EB and adjusted the figure slightly to avoid any misunderstanding.

After Eq 4 the discussion suggests that for ϕ << 1 the threshold energy is much greater than kBT, but that’s hard to imagine given that the logarithmic dependence of the latter on the former. Also in Figure 2d it seems that the threshold energy is about 8 kBT. Clearly this is not a big deal, just thought the authors might want to revise the language.

Thank you. We now clarify the sentence using the representative values of ϕ and Eth after Equation 4.

Right after Figure 2 there is a discussion of the different parameters that the authors vary. I suggest having a figure that illustrates these parameters (possibly in Figure 1b) to make it easier to follow the discussion.

We have added explanations of the relevant parameters in Figure 1 for clarity.

When discussing the dynamics of search the result stated is that the search time is minimum for a specific value of R. I think it would be useful to translate this into a TF concentration. Also, if R represents the radius of the cells nucleus 1/6 um is almost an order of magnitude smaller than the size of a typical nucleus. Is this a worry? Either way some clarification of this number would be helpful.

Thank you for the suggestion. As noted later in this section, we have translated R into an equivalent TF concentration, and we clarify that we assume the scaling of the minimum search time remains unchanged when extrapolated to the size of a typical nucleus.

There is a comment regarding the role of the DNA persistence length and how it was not accounted for. It would be helpful if the authors could add a sentence or two explains how a folded DNA conformation, as is the case in the nucleus, would affect their calculation. (So that the reader gets an idea without having to get into the details described in the Supplement).

Thank you. We have revised the sentence to: “We have verified that reducing the DNA persistence length, which promotes increased DNA coiling, results in only a modest increase in mean search time. Even under extreme coiling conditions, the increase remains below 30% of the baseline value, as detailed in Supplementary S9.”.

Version 2

Abstract

Abstract

Transcription Factors (TFs) are proteins crucial for regulating gene expression. Effective regulation requires the TFs to rapidly bind to their correct target, enabling the cell to respond efficiently to stimuli such as nutrient availability or the presence of toxins. However, the search process is hindered by slow diffusive movement and the presence of ‘false’ targets - DNA segments that are similar to the true target. In eukaryotic cells, most TFs contain an Intrinsically Disordered Region (IDR), which is commonly assumed to behave as a long, flexible polymeric tail composed of hundreds of amino acids. Recent experimental findings indicate that the IDR of certain TFs plays a pivotal role in the search process. However, the principles underlying the IDR’s role remain unclear. Here, we reveal key design principles of the IDR related to TF binding affinity and search time. Our results demonstrate that the IDR significantly enhances both of these aspects. Furthermore, our model shows good agreement with experimental results, and we propose further experiments to validate the model’s predictions.

Introduction

Transcription Factors (TFs) are fundamental proteins that regulate gene expression: they bind to specific DNA sequences to control gene transcription [1]; they respond adaptively to various cellular signals, thereby modulating gene expression [2, 3]; they guide processes such as cell differentiation and development, and their malfunctions are often linked to disease onset [4, 5]. The principles of gene regulation were first formulated in the 1960s through pioneering studies of the E. coli Lac operon [6], establishing that the regulation mechanism is via the binding of a TF to the regulatory DNA regions associated with the genes, the TF’s target. Effective regulation requires TFs to have sufficient binding affinity and a binding rate comparable to cellular processes. Despite these decades of research, the factors determining the binding strength and search efficiency of eukaryotic TFs remain incompletely understood [7, 8]. In eukaryotic cells, besides having a DNA-binding domain (DBD), approximately 80% of TFs also have one or more intrinsically disordered regions (IDRs), which lack a stable 3D structure and can be assumed as flexible, unstructured polymers, hundreds of amino acids (AAs) in length [9–12]. Eukaryotic DBDs often exhibit weaker affinity to their recognition sequences compared to their bacterial counterparts, indicating that IDRs are necessary for achieving stable binding [13]. However, IDRs are challenging to study because their functions can be maintained despite rapid sequence divergence, complicating traditional sequence-based analyses [14, 15]. Molecular dynamics simulations suggest that IDRs play a significant role in promoting target recognition [16], by increasing the area of effective interaction between the TF and the DNA. Upon binding, the search problem becomes effectively a one-dimensional diffusion process [17]. Recent in vivo experiments done in yeast also suggest that IDRs could take a key part in guiding TFs to the DNA regions containing their targets [18, 19]. Truncation studies of IDRs in various TFs have demonstrated that removing these regions often results in reduced binding affinity, suggesting they contribute significantly to stabilizing TF-DNA interactions [18]. Different mechanisms have been proposed in the literature to explain IDR contributions: some suggest non-specific electrostatic interactions [20], others propose phase separation that concentrates TFs near their targets [21], while a third view suggests specific IDR-DNA interactions analogous to those of structured domains [22].

The problem of TF search has been widely studied, theoretically, albeit with the majority of works focusing on parameter regimes relevant for bacteria. One important mechanism is that of facilitated diffusion, in which the TF performs 1D diffusion along the DNA and at any given moment can fall off and perform 3D diffusion until reattaching to the DNA (a 3D excursion). The 1D diffusion along the DNA and the 3D excursions alternate until the TF finds its target. This mechanism is well characterized theoretically [23–29] and supported by experiments in E. coli [30]. However, due to the significant structural differences, including larger genome size and chromatin structure, the facilitated diffusion mechanism cannot explain the fast search times observed in eukaryotes, and a qualitatively different mechanism is required [7, 19, 31–33].

Here, we attempt to address the fundamental question of how IDRs can enhance the binding affinity of the TFs and speed up their search process, and to discover the associated design principles in eukaryotes. The problem of binding affinity (binding probability) exhibits the classical trade-off between energy and entropy: on one hand, the binding of IDR to the DNA is energetically favorable. On the other hand, this binding leads to constraints on the positions of the IDR (pinned to the DNA at the points of attachment) which leads to a reduction in the entropy associated with the possible polymer conformations - hence an increase in the free energy of the system. Therefore the effects of the disordered tail on affinity are non-trivial, leading to physical problems reminiscent of those associated with the well-studied problem of DNA unzipping [34]. We find that the binding probability increases dramatically for TFs having IDRs for a broad parameter regime. The optimal search process consists of one round of 3D-1D search, where at first the TF performs 3D diffusion and then binds the antenna and performs 1D diffusion until finding its target, rarely unbinding for another 3D excursion. This process is presented in Fig. 1(a, c). We obtain an expression for the search time that shows good agreement with our numerical results.

Model illustration of a TF with the IDR and its search process.

a, Illustration of a TF locating its target site (highlighted in orange) on the DNA strand (depicted as a green curve). Surrounding the target is a region of length L, where the TF’s IDR interacts with the DNA. The TF performs 3D diffusion until it encounters and binds to this “antenna” region. b, Illustration of a TF that includes an IDR composed of AAs. On the right: a model of the TF, a polymer chain, comprising of a single binding site on the DNA Binding Domain (DBD), and multiple binding sites on the IDR (also known as short linear motifs [8]). c, Once bound to the antenna, the TF performs effective 1D diffusion until it reaches its target. The 1D diffusion is via the binding and unbinding of sites along the IDR, a process we coin “octopusing”. V1 is the targets’ volume, d the separation, and EB the binding energy per binding site, where each site corresponds to a short linear motif.

Results

Binding probability of the TF from the equilibrium approach

The experiment described in Ref. [18] studied the binding affinity of wild-type and mutated TFs to their corresponding targets. They found that shortening the IDR length results in weaker affinity. To qualitatively model the binding affinity of a TF to its target, we utilize a well-known physical model of polymers, the Rouse Model [35, 36] (used in the subsequent dynamical section as well). See Materials and Methods. We note that our conclusions do not hinge on this choice, and our results hold also when using other polymer models (see Supplementary S1). Since the IDR is thought to recognize a relatively short flanking DNA sequence of the DBD target [18], we model a DNA segment as a straight-line “antenna” region [26, 37–39], where the IDR targets reside. We assume that the interaction between TF and DNA outside the antenna region is negligible. Note that the interaction between IDR and DNA is a specific interaction, with genomic localization determined by multivalent interactions mediated by hydrophobic residues [40, 41].

To gain an intuition into the binding probability of TF, we first study a TF with a single IDR site. From here on, energies will be measured in units of kBT, to simplify our notation. We assume that the IDR has only one corresponding target. There are four possible states: neither the DBD nor the IDR are bound, the IDR is bound, the DBD is bound, and both are bound, which correspond to Pfree, PIDR, PDBD , and Pboth, respectively.

The probability that the TF is bound at the DBD target is the sum of PDBD and Pboth. In thermal equilibrium, each of the 4 terms is given by the Boltzmann distribution, which can be solved analytically by integrating-out the additional degrees of freedom associated with the polymer (Supplementary S2). To better understand the associated design principles, it will be helpful to compare this probability with that of a “simple” TF (i.e., without the IDR), which we denote by Psimple. We find that: PTFPsimple≈(1+P), (1)

where P is an enhancement factor, governing the degree to which the IDR improved the binding affinity. Importantly, we find that: P(EB,d,l0)=eEBV1f(d,l0), (2)

where f(d,l0)=(32πl02)3/2exp(−3d22l02) , is the probability density, with units of inverse volume, V1 is the targets’ volume, on the order of 1-10bp3, as shown in Fig. 1c, -EB is the binding energy for per IDR target (EB is always positive), d is the distance between the DBD target and the IDR target, and l0 is the distance between the DBD and the IDR binding site on the TF (l0 and d are also the distances for neighboring IDR targets and neighboring IDR sites, respectively, in the case of multiple IDR targets and sites). This shows that the presence of the IDR always increases the binding affinity since 𝒫 is always positive. However, the magnitude of enhancement can be negligible: For instance, even if EB is large, the term f(d, l0), associated with the entropic penalty (because the end-to-end distance is fixed, reducing the degrees of freedom compared to a free polymer), could lead to small values of 𝒫. The exponential decay of 𝒫 with d2 (through the term f(d, l0)), indicates that placing the IDR targets close to the DBD target is preferable. However, due to physical limitations, d cannot be arbitrarily small and in reality will have a lower bound. Once d is fixed, 𝒫 is maximal at l0 = d, as can be gleaned from equation (2). This condition minimizes the free energy penalty associated with the entropic contribution of the term f(d, l0). As we shall see, this design principle is also intact when multiple IDR sites and targets are involved, which is the scenario we discuss next.

A priori, it is unclear whether having more binding sites (nb > 1) on the IDR or more targets (nt > 1) on the DNA will be beneficial or not. We therefore sought to investigate how the binding affinity depends on these parameters, as well as the binding energy and positions of the targets and binding sites. This generalization requires taking into account all the possible number of bindings, n (ranging from 0 to min (nb, nt)), and all possibilities for binding orientations for each n, (nbn)(ntn)n! distinct configurations in total. For each of them, we may analytically integrate-out the polymer degrees of freedom and calculate explicitly the contribution to the partition function (Supplementary S2). Summing these terms together, numerically, allows us to explore how the different parameters affect the binding specificity, which we describe next. To characterize the binding affinity, we resort again to the ratio Q = PTF/Psimple, measuring the effectiveness of the binding in comparison to the case with no IDRs. The length of TFs in our considerations, which varies with nb, is comparable to that in experiments [18].

We find the design principle: l0 = d, identical to what we showed for the case of a single IDR binding site. Specifically, Q is maximal at l0 ~ d for all combinations of nb and nt. Fig. 2a displays this principle for several combinations of nb and nt. All combinations are provided in Supplementary Fig. S3. This is plausible since for the IDR to be bound to the DNA at multiple places without paying a large entropic penalty, the distance between target sites should be compatible with the distance between binding sites. We therefore set l0 = d and search for other design principles. Fig. 2b shows a heatmap of Q varying with nb and nt. At large nt, as nb increases, the value of PTF exhibits an initial substantial increase, followed by a gradual rise until it reaches saturation (shown explicitly in Supplementary Fig. S4). A similar behavior is also observed when varying the value of nt. (When nt exceeds nb, the contribution from the states where the IDR is bound but the DBD is unbound increases with increasing nt, which slightly lowers PTF; see Supplementary Fig. S4.) In fact, the heatmap shown in Fig. 2b shows that the value of PTF is governed, approximately, by the minimum of nb and nt (this is reflected in the nearly symmetrical “L-shapes” of the contours). This places strong constraints on the optimization, and suggests that there will be little advantage in increasing the value of either nb or nt beyond the point nb = nt, revealing another design principle (under the assumption that it is preferable to have smaller values of nb and nt for a given level of affinity, in order to minimize the cellular costs involved). We verify this principle holds at additional values of EB. Fig. 2c shows that the contours corresponding to PTF = 0.9 at varying values of EB all take an approximate symmetrical L-shape, thus supporting the design principle nb = nt. For a given EB, we denote the optimal values as nb*=nt* , and show them as hexagrams in the plot. We further find that nb* decreases with increasing EB, as displayed in the inset. These relationships also hold at PTF = 0.5 (Supplementary S5).

We examine the variation of Q with respect to EB while keeping nb = nt constant. In Fig. 2d, as the binding energy EB increases, Q increases from a value of ~ 1 to its saturation value, 1/Psimple, in a switch-like manner: i.e., there is a characteristic energy scale, Eth. To estimate Eth, when Q ≪ 1/Psimple (i.e., PTF 1), we approximate the expression of Q as: Q≈1+∑n=1nb∑orn∏i=1nP(EB,rorn,i,lorn,i), (3)

where lorn, i and rorn, i denote the segment length between two adjacent bound sites and the distance associated with their respective targets, respectively, and “orn” represents the orientation of the given configuration (Supplementary S2). Equation (3) reduces to equation (1) when nt = nb = 1. To substantially increase Q, the configuration that contributes the most to Q, would satisfy (𝒫(EB, d, d))nb ≪ 1. Thus, the characteristic energy Eth could be estimated using the formula 𝒫(Eth, d, d) = 1, resulting in: Eth=32[ln(2π3ϕ2)+1], (4)

where ϕ=V11/3/d represents the proportion of IDR targets within the antenna. Equation (4) is a direct consequence of the energy-entropy trade-off. Increasing ϕ leads to a decrease in entropy loss due to binding the IDR, subsequently reducing the demands on binding energy. Given that ϕ2 = 0.02, Eth is about 8.5kBT. The vertical dashed line in Fig. 2d, obtained from equation (4), effectively captures the region of significant increase for large values of nb. Hence, we obtain the third design principle: EB > Eth, which ensures the significant binding advantage of TFs with IDRs.

Design principles across various parameters: l0/d, nb, nt, and EB.

a, Plots of Q = PTF/Psimple varying with l0/d are shown for several combinations of nt and nb for typical values of the relevant quantities: the nucleus volume is 1μm3, EDBD = 15, EB = 10, V1 = (0.34nm)3, and d=50⋅0.34nm . b, A heatmap of Q vs. nb and nt at EB = 10 and d = l0. c, Contour lines at different EB at PTF = 0.9. Blue hexagrams represent the design principle nb*=nt* , with the black dashed line as a visual guide. The inset shows the dependence of nb* on energy. d, Q for varying EB at d = l0 for several nb = nt. The vertical line represents Eth as determined by equation (4), the solid curves illustrate the approximation of Q obtained from equation (3), and the horizontal line indicates where PTF = 1.

So far, we have found that the binding design principles for TFs with IDRs are as follows: (i) target distance d should be as small as possible, (ii) IDR segment binding distance l0 should be comparable to d, (iii) it is preferable to have fewer binding sites nb* and binding targets nt* as the binding strength EB increases, with nb*~nt* , and (iv) EB should be larger than a threshold that solely depends on the proportion ϕ of IDR targets within the antenna, as indicated by Eq. (4). In our estimation, with ϕ2 = 0.02, we obtain Eth ≈ 8.5kBT. The above principles have also been validated in scenarios involving Poisson statistics of binding site distances and target distances, thus indicating their robust applicability (Supplementary Fig. S6).

Search time of the TF from the dynamics

In this section, we study the design principles of the TF search time, a dynamic process. Here, we confine the TF with ñ = nb+1 sites (nb IDR sites and one DBD site) to a sphere of radius R, the cell nucleus, and use over-damped coarse-grained molecular dynamics (MD) simulations. The sites and targets interact via a short-range interaction. The DBD target with a radius a is placed at the sphere’s center. See the sketch in Fig. 1a. Once the DBD finds its target, the search ends. We denote the mean total search time (i.e., the mean first passage time, MFPT), as ttotal. We estimate it by averaging 500 simulations, each starting from an equilibrium configuration of the TF at a random position in the nucleus. Simulation details are provided in the Materials and Methods section.

By conducting comprehensive simulations of ttotal across various parameter values of L, EB, and ñ, we were able to identify its minimum value, which is approximately at L* = 300bp, EB* = 11, and ñ* = 4 for R = 500bp ≈ 1/6μm. Here, one base pair (bp) is equal to 0.34nm. To speed up the simulations, we utilized several computational tricks, including adaptive time steps and treating the TF as a point particle when it is far from the antenna (see Materials and Methods). Despite of this, the simulations require 105 CPU hours due to the IDR trapped in the binding targets on the antenna, which results in the slow dynamics as found in the previous simulation works [16, 42]. The probability distribution function (pdf) of individual search times exponentially decays with a typical time equal to ttotal, see inset of Fig. 3. This indicates that ttotal provides a good characterization of the search time.

We find that ttotal varies non-monotonically with L at fixed EB* and ñ* as shown in Fig. 3. Once a TF attaches to one of its targets, it will most likely diffuse along the antenna until reaching the DBD target (Supplementary S7). It means that the optimal search process is a single round of 3D diffusion followed by 1D diffusion along the antenna. In Supplementary S8, by analytically calculating the mean total search time using a coarse-grained model, we find that a single round of 3D-1D search is typically the case in the optimal search process. Note that our model differs profoundly from the conventional facilitated diffusion (FD) model in bacteria: In the latter, the TF is assumed to bind non-specifically to the DNA. Within our model, the antenna length L can be tuned by changing the placement of the IDR targets on the DNA, making it an adjustable parameter rather than the entire genome length. As opposed to the FD model, which requires numerous rounds of DNA binding-unbinding to achieve minimal search time, within our model, the minimal search time is achieved for a single round of 3D and 1D search.

Interestingly, when we observe the motion of a TF along the antenna, IDR sites behave like “tentacles”, as illustrated in Fig. 1c, constantly binding and unbinding from the DNA. We refer to this 1D walk as “octopusing”, inspired by the famous “reptation” motion in entangled polymers suggested by de Gennes [43] (see Supplementary Video). We note that previous numerical studies [16, 17] used comprehensive, coarse-grained, MD simulations to demonstrate that including heterogeneously distributed binding sites on the IDR could facilitate intersegmental transfer between different regions of the DNA, which is reminiscent of our octopusing scenario.

The mean 3D search time, t3D, and the mean 1D search time, t1D are also shown in Fig. 3. To evaluate t3D, we take advantage of several insights: first, if the TF gyration radius, rp=n˜/6l0 , is comparable or larger than the distance between targets, d, we may consider the chain of targets as, effectively, a continuous antenna. Making the targets denser would hardly accelerate the search time, reminiscent of the well-known chemoreception problem where high efficiency is achieved even with low receptor coverage [44]. Second, when the distance between the TF and the antenna is smaller than rp, one of the binding sites on the TF would almost surely attach to the targets. This allows us to approximately map the problem to the MFPT of a particle hitting an extremely thin ellipsoid with long-axis of length L and radius rp, the solution of which is ln(L/rp)/L [45]. We find that this scaling captures well the dependence of t3D on the parameters, with a prefactor α ~ O(1). For the 1D search from a single particle picture, since the TF is unlikely to detach once it attaches, it performs a 1D random walk. So, taken together, the mean total search time is ttotal=t3D+t1D≈αn˜R3DLln(L/rp)+βL2D, (5)

where D is the 3D diffusion coefficient of one site. For a TF with ñ sites, its 3D diffusion coefficient, D3, is D/ñ. α and β are fitting parameters. The semi-analytical formulas of t1D, t3D, and ttotal, denoted by the solid lines in Fig. 3, agree well with the simulation results for α = 0.5 and β = 14.5. The theoretical result for the time spent in 3D was originally calculated for a point searcher [45]. In that case, a value of α = 2/3 was obtained. β is larger at larger EB (see Supplementary S7).

Mean search time varying with the antenna length L.

ttotal, t3D and t1D vs. L at EB* and ñ*. The solid curves correspond to ttotal, t3D and t1D in equation (5). Inset: probability density function (pdf) exponentially decays at large t. pdf ≈ exp(−t/ttotal)/ttotal (black solid line). The unit t0 represents the time for one AA to diffuse 1bp. a = 0.5bp, d = 10bp, and l0 = 5bp.

We also examine the dependence of ttotal on the key parameters EB and ñ, in the vicinity of the optimal parameters which minimize the total search time (Supplementary S7). We find that the affinity of TFs to the antenna decreases rapidly with decreasing the IDR length [18] and EB. Consequently, when ñ < ñ* or EB>EB* , multiple search rounds are needed to reach the DBD target, resulting in a significant escalation of ttotal. When ñ>ñ*, ttotal, t3D, and t1D agree with equation (5). When EB>EB* , since t3D is independent of EB, the variation in ttotal stems from t1D which exponentially increases with EB.

The minimal search time for a simple TF is tsimple = R3 /(3Da) [45]. For the TF with an IDR, by minimizing the expression in equation (5), we find that the optimal length approximately scales as L* ∞ R, and therefore the minimal search time, tmin, is approximately proportional to R2. At R = 500 bp ≈ 1/6μm, the search time is more than ten times shorter than a simple TF, despite the latter’s diffusion coefficient being several times larger. Based on our analytic result, we expect ~ 50 fold enhancement for the experimental parameters of yeast (note that we assume the scaling tmin ~ R2 still holds in this regime; this formula is approximate since it does not take into account the persistence length of DNA [46], which is several-fold shorter than the optimal antenna length L*. We have verified that reducing the DNA persistence length, which promotes increased DNA coiling, results in only a modest increase in mean search time. Even under extreme coiling conditions, the increase remains below 30% of the baseline value, as detailed in Supplementary S9. Converting tmin into a second-order on-rate yields approximately 108 — 109 M−1 s−1, assuming D ≈ 10−13m2/s and a TF copy number of 100. We also verified that the 3D diffusion coefficient slightly changes under a complex DNA configuration, where a point-searcher TF non-specifically interacts with the entire DNA, indicating the robustness of our optimal search process (see Supplementary S10).

Next, we characterize the dynamics of the octopusing walk (Supplementary S11). During this process, different IDR binding sites constantly bind and unbind from the antenna, leading to an effective 1D diffusion. A priori, one may think that the TF faces conflicting demands: it has to rapidly diffuse along the DNA while also maintaining strong enough binding to prevent detaching from the DNA too soon. Indeed, within the context of TF search in bacteria, this is known as the speed-stability paradox [25]. Importantly, the simple picture of the octopusing walk explains an elegant mechanism to bypass this paradox, achieving a high diffusion coefficient with a low rate of detaching from the DNA. This comes about due to the compensatory nature of the dynamics: there is, in effect, a feedback mechanism whereby if only a few sites are bound to the DNA, the on-rate increases, and similarly, if too many are bound (thus prohibiting the 1D diffusion) the off-rate increases. Remarkably, this mechanism remains highly effective even when the typical number of binding sites is as low as two. Further details are provided in Supplementary S11.

Comparing with experimental results

In Ref. [18], the binding probability of the TF Msn2 in yeast was measured, where the native IDR was shortened by truncating an increasingly large portion of it. The results indicate that binding probability decreases with increasing truncation of the IDR, as illustrated in Fig. 4(a). We utilized our model to calculate the relative binding probability compared to PTF at a zero truncating length. Using the DBD binding energy EDBD as our only fitting parameter, our model achieves quantitative agreement with the experimental data. In Fig. 4(b), we estimate the search time as a function of L for R = 1μm and D0 = 10−10m2/s for a single AA. The estimated timescales are comparable to empirical results, which are 1 to 10 seconds [19, 47], as indicated by the shaded area. We have also estimated the on-rate kon and off-rate koff of the TF as they vary with IDR length, as shown in Fig. 4(c). In the kinetic proofreading mechanism [48], specificity is associated with the off-rate, but recent studies [49] suggest that in other scenarios, specificity is encoded in the on-rate. Therefore, it is important to explore how specificity is encoded within this mechanism. Our results indicate that the variation in the dissociation constant koff/kon is dominated by koff, suggesting that specificity is encoded in the off-rate. The magnitudes and variations of these rates could be directly compared to experimental results in future studies.

a, Relative binding probability varies with the truncation length of the IDR, which quantitatively agrees with the experimental data in Ref. [18], where we only vary EDBD (binding energy of DBD) to ensure that the maximal value reaches 1 at zero truncation length. Antenna length L = 1000nm, EDBD =22 and EB = 11. Other parameters are set the same as in Fig. 2. b, The search time estimated by our model (from Eq. (5) and divided by a TF copy number of 100) is quantitatively comparable with the experimental results [47], as guided by the shaded area. c, The on-rate kon = ttotal/Vc and off-rate koff = (1 − PTF)/(PTFttotal) varying with the IDR length, indicating that kD = koff/kon ranges from 0.01 to 10 nM, with its variation primarily dominated by the off-rate.

Conclusion and discussion

In summary, we addressed the design principles of TFs containing IDRs, with respect to binding affinity and search time. We find that both characteristics are significantly enhanced compared to simple TFs lacking IDRs.

The simplified model we presented provides general design principles for the search process and testable predictions. One possible experiment that could shed light on one of our key assumptions is “scrambling” the DNA (randomizing the order of nucleotides). Our results depend greatly on our assumption that there exists a region on the DNA, close to the DBD target, that contains targets for the IDR binding sites. It is possible to shuffle, or even remove, regions of different lengths around the DBD target and to measure the resulting changes in binding affinity. According to our results, we would expect that shuffling a region smaller in length than the IDR should have minimal effect on the binding affinity and search time, but that shuffling a region longer than the IDR should have a noticeable effect.

Another useful test of our model would be to probe the dynamics of the TF search process by varying the waiting time before TF-DNA interactions are measured: Within the ChEC-seq technique [50] employed in Ref. [18], an enzyme capable of cutting the DNA is attached to the TF, and activated at will (using an external calcium signal). Sequencing then informs us of the places on the DNA at which the TF was bound at the point of activation. Measuring the binding probabilities at various waiting times — defined as the time between adding calcium ions and adding transcription factors, ranging from seconds to an hour, would inform us both of the TF search dynamics as well as the equilibrium properties: Shorter time scales would capture the dynamics of the search process, while longer time scales would reveal the equilibrium properties of the binding. At short waiting times (within several seconds), we would expect the binding probability to increase with waiting times. By analyzing a normalized binding probability curve as a function of the waiting time t, we could fit the curve using the function 1-exp(−t/tsearch). From this curve, we can infer the search time. Our results suggest that the presence of the IDR in a TF leads to increased binding probabilities over various waiting times and a reduced waiting time to reach a higher saturation binding probability. Furthermore, as shown in Fig. 4 obtained from our model, kon can be determined from the measurements of tsearch. Using the binding probability, which is associated with kD = koff/kon, koff can also be calculated.

While our work primarily focused on coarse-grained modeling of the search process and the hypothesized octopusing dynamics, higher-resolution molecular dynamics simulations could provide more detailed molecular insights and further validate our proposed design principles [16, 17]. Since these principles are grounded in fundamental physical concepts, such as the energy-entropy trade-off, they are likely to be robust across diverse molecular systems. We anticipate their applicability to other biophysical contexts, such as nonspecific RNA polymerase binding [51] and multivalent antibody binding [52], where the associated dynamic processes would also be of significant interest for future research. Beyond our octopusing mechanism, other mechanisms may also contribute to the search and binding process. IDRs could facilitate complex formation and cooperativity between multiple transcription factors, as suggested by studies showing co-localization at common target promoters [53]. IDRs might also enhance efficiency by directing transcription factors to specific nuclear compartments or biomolecular condensates, thereby reducing the effective search space and time [21, 54, 55]. These complementary mechanisms [7, 8, 56] suggest that our findings represent one component of a broader regulatory framework governing transcriptional control in eukaryotic cells.

Materials and methods

Model

The Rouse Model describes a polymer as a chain of point masses connected by springs and characterized by a statistical segment length l0 resulting from the rigidity of springs k and the thermal noise level kBT, l02 = 3kBT/k [35, 36]. We coarse-grain the TF by representing it with a few binding sites for the IDR, and an additional site for the DBD, as illustrated in Fig. 1b. The targets are assumed to be positioned at equal spacing, with one target for the DBD and a few targets for the IDR, corresponding to the orange and black squares in Fig. 1c. We have also verified that relaxing this assumption, and using randomly positioned targets, does not affect our results (Supplementary S6). In the Rouse Model, the 3D probability density describing the relative distance in the equilibrium state, denoted as r, between any two sites can be derived through Gaussian integration [36], leading to: f(r,l)=(32πl2)32exp(−3r22l2), (6)

where l=nsegl0 and nseg is the number of segments between the two sites.

Dynamical simulations

Each binding site of the TF at position ri follows over-damped dynamics, which reads: Δri=DkBTf({ri})Δt+2DΔtξ, (7)

where the random variable ξ follows the standard normal distribution. D is the diffusion coefficient of the coarsegrained binding site. The force acting on each site is f ({ri}) = −∂Utotal/∂ri, and Utotal=∑i=0nb−112k(ri+1−ri)2+∑i,jEBexp[−(ri−Rj)22wd2], (8)

where Rj are the positions of the targets. wd = 1bp is the typical width of the short-range interaction between sites and their corresponding targets. The typical distance l0 between coarse-grained binding sites is 5bp. The stiffness of the springs is thus 3kBT/ l02 = 3kBT/25bp2. To speed up the simulations, when the sites’ minimal distance to the antenna is larger than the threshold 10bp, the TF is regarded as a rigid body, and all sites move in sync with their center of mass. We verified that the resulting search time is insensitive to the value of this threshold. When the sites’ minimal distance to the antenna is less than 10bp, we consider the excluded volume effect to ensure that different binding sites are not trapped at the same target. Specifically, we use the WCA interaction [57]: UWCA={σ12|ri−rj|12−2σ6|ri−rj|6+1|ri−rj|<σ0,|ri−rj|≥σ (9)

where σ = 1bp. In this region, we also adopted an adaptive time step: Δt/(bp2/D) = min{1,kBTmax|fi|bp}×0.01 . Note that this choice ensures that the displacement of the sites at every step is much smaller than the binding site size wd. Due to the IDR trapped in the binding targets on the antenna, which results in slow dynamics, the simulation time becomes significantly longer, as found in previous simulation studies [16, 42]. This slowdown is similar to the behavior observed in glass-forming liquids [58]. As a result, the individual simulation time steps range from 109 to 1011.